Fast sparse period estimation

### 2014-03-28 paper.tex
minor
added newext build

The problem of estimating the period of a point process from observations that are both sparse and noisy is considered.  By sparse it is meant that only a potentially small unknown subset of the process is observed.  By noisy it is meant that the subset that is observed, is observed with some error, or noise.  Existing accurate algorithms for estimating the period require $O(N^2)$ operations where $N$ is the number of observations.  By quantising the observations we produce an estimator that requires only $O(N\log N)$ operations by use of the chirp z-transform or the fast Fourier transform.  The quantisation has the adverse affect of decreasing the accuracy of the estimator.  This is investigated by Monte-Carlo simulation.  The simulations indicate that significant computational savings are possible with negligible loss in statistical accuracy.

@@

The maximiser $\widetilde{f}^\prime$ can be computed efficiently using the chirp z-transform.  Let $b_i$ be the number of integers from $\ell_1,\dots,\ell_N$ that are equal to $i$.  That is,
\[
b_i= \#\{ n \mid \ell_n = i, n = 1,\dots, N\}
\]
where $\#$ indicates the number of elements in a set.  The integers $\ell_1,\dots,\ell_N$ are in ascending order because $y_1,\dots,y_N$ are in ascending order.  So $b_i = 0$ if $i < \ell_1$ or $i > \ell_N$.  Now
\[
I_z(f) = \babs{ \sum_{i = \ell_1}^{\ell_N} e^{ -2\pi j f i/q} b_i }^2 = \babs{ \sum_{m = 0}^{\ell_N-\ell_1} e^{ -2\pi j f m/q} b_{m+\ell_1} }^2.
\]
We want to compute $I_z(f)$ for all $f$ in the grid $G$.  Putting
\[
v_m = e^{ -2\pi j f_{\text{min}}m/q} b_{m+\ell_1}
\]
we have
\[
I_z(f_{\text{min}} + Dk) = \babs{ \sum_{m = 0}^{\ell_N-\ell_1} e^{ -2\pi j Dk m/q} v_m }^2 = \abs{V_k}^2
\]   
where, using the notation of Rabiner~et.~al.~\cite[eq.~(10)]{Rabiner1969},
\[
V_k = \sum_{m = 0}^{\ell_N-\ell_1} e^{ -2\pi j Dk m/q} v_m = \sum_{m=0}^{\ell_N-\ell_1} A^{-m} W^{km} v_m
\]
where $A = 1$ and $W = e^{ -2\pi j D/q}$.  All of $V_0,\dots,V_{K-1}$ can be efficiently computed by the chirp z-transform~\cite{Rabiner1969}.  Putting
\[
\hat{k} = \arg\max_{k = 0,1,\dots,K-1} \abs{ V_k }^2
\]
we have
\[ 
\widetilde{f}^\prime = f_{\text{min}} + D\hat{k} = \arg\max_{f \in G} I_z(f).
\]
The frequency $\widetilde{f}^\prime$ can be used to initialize an iterative procedure to maximise $I_y$.  The procedure will converge to $\hat{f}$ if $\widetilde{f}^\prime$ is sufficiently close to $\hat{f}$.

@@

@@

%      \addplot[mark=none,color=black,dashed] table {code/data/NormalisedLLSCLTN30};

@@

 %     \addplot[mark=none,color=black,dashed] table {code/data/NormalisedLLSCLTN1200};

@@

     \legend{Periodogram theory, $q=\tfrac{3}{2}f_{\text{max}}$, $q=4 f_{\text{max}}$, Periodogram, Least squares, SLS2-new}

@@

@@

 %     \addplot[mark=none,color=black,dashed] table {code/data/NormalisedLLSCLTN1100};

@@

%Currently, as is common An open theoretical question is a rigorously

\small
%\bibliography{zotero}
\bibliography{bib}

>>>>

The problem of estimating the period of a point process from observations that are both sparse and noisy is considered.  By sparse it is meant that only a potentially small unknown subset of the process is observed.  By noisy it is meant that the subset that is observed, is observed with some error, or noise.  Existing accurate algorithms for estimating the period require $O(N^2)$ operations where $N$ is the number of observations.  By quantising the observations we produce an estimator that requires only $O(N\log N)$ operations by use of the chirp z-transform or the fast Fourier transform.  The quantisation has the adverse effect of decreasing the accuracy of the estimator.  This is investigated by Monte-Carlo simulation.  The simulations indicate that significant computational savings are possible with negligible loss in statistical accuracy.

@@

The maximiser $\widetilde{f}^\prime$ can be computed efficiently using the chirp z-transform.  Let $b_i$ be the number of integers from $\ell_1,\dots,\ell_N$ that are equal to $i$.  That is,
\[
b_i= \#\{ n \mid \ell_n = i, n = 1,\dots, N\}
\]
where $\#$ indicates the number of elements in a set.  The integers $\ell_1,\dots,\ell_N$ are in ascending order because $y_1,\dots,y_N$ are in ascending order.  So $b_i = 0$ if $i < \ell_1$ or $i > \ell_N$.  Now
\[
I_z(f) = \babs{ \sum_{i = \ell_1}^{\ell_N} e^{ -2\pi j f i/q} b_i }^2 = \babs{ \sum_{m = 0}^{\ell_N-\ell_1} e^{ -2\pi j f m/q} b_{m+\ell_1} }^2.
\]
We want to compute $I_z(f)$ for all $f$ in the grid $G$.  Putting
\[
v_m = e^{ -2\pi j f_{\text{min}}m/q} b_{m+\ell_1}
\]
we have
\[
I_z(f_{\text{min}} + Dk) = \babs{ \sum_{m = 0}^{\ell_N-\ell_1} e^{ -2\pi j Dk m/q} v_m }^2 = \abs{V_k}^2
\]   
where, using the notation of Rabiner~et.~al.~\cite[eq.~(10)]{Rabiner1969},
\[
V_k = \sum_{m = 0}^{\ell_N-\ell_1} e^{ -2\pi j Dk m/q} v_m = \sum_{m=0}^{\ell_N-\ell_1} A^{-m} W^{km} v_m
\]
where $A = 1$ and $W = e^{ -2\pi j D/q}$.  The $V_0,\dots,V_{K-1}$ can be efficiently computed by the chirp z-transform~\cite{Rabiner1969}.  Putting
\[
\hat{k} = \arg\max_{k = 0,1,\dots,K-1} \abs{ V_k }^2
\]
we have
\[ 
\widetilde{f}^\prime = f_{\text{min}} + D\hat{k} = \arg\max_{f \in G} I_z(f).
\]
The frequency $\widetilde{f}^\prime$ can be used to initialize an iterative procedure to maximise $I_y$.  The procedure will converge to $\hat{f}$ if $\widetilde{f}^\prime$ is sufficiently close to $\hat{f}$.

@@

      \addplot[mark=none,color=black,dashed] table {code/data/NormalisedLLSCLTN30geom3};

@@

@@

      \addplot[mark=none,color=black,dashed] table {code/data/NormalisedLLSCLTN1200geom3};

@@

     \legend{Periodogram theory, Least squares theory, $q=\tfrac{3}{2}f_{\text{max}}$, $q=4 f_{\text{max}}$, Periodogram, Least squares, SLS2-new}

@@

      \addplot[mark=none,color=black,dashed] table {code/data/NormalisedLLSCLTN30geom15};

@@

      \addplot[mark=none,color=black,dashed] table {code/data/NormalisedLLSCLTN1200geom15};

@@

\small
\bibliography{bib}
